Distortion corrected kernel density estimate on Riemannian manifolds

### 2023-06-01 jcgs/kderm_jcgs.Rmd
Add jcgs version for kderm, the latest version

\documentclass[11pt,a4paper,]{article}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Distortion corrected kernel density estimate on Riemannian manifolds},
            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{Distortion corrected kernel density estimate on Riemannian manifolds}

%% MONASH STUFF

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% FONT
\RequirePackage{bera}
\RequirePackage{mathpazo}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% MATHS
\RequirePackage{bm,amsmath}
\allowdisplaybreaks

%% GRAPHICS
\RequirePackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.8}

%\RequirePackage[section]{placeins}

%% SECTION TITLES
\RequirePackage[compact,sf,bf]{titlesec}
\titleformat{\section}[block]
  {\fontsize{15}{17}\bfseries\sffamily}
  {\thesection}
  {0.4em}{}
\titleformat{\subsection}[block]
  {\fontsize{12}{14}\bfseries\sffamily}
  {\thesubsection}
  {0.4em}{}
\titlespacing{\section}{0pt}{*5}{*1}
\titlespacing{\subsection}{0pt}{*2}{*0.2}

%% TITLE PAGE
\def\Date{\number\day}
\def\Month{\ifcase\month\or
 January\or February\or March\or April\or May\or June\or
 July\or August\or September\or October\or November\or December\fi}
\def\Year{\number\year}

\makeatletter
\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
\def\nojel{\def\showjel{}}
\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
\def\cover{{\sffamily\setcounter{page}{0}
        \thispagestyle{empty}
        \placefig{2}{1.5}{width=5cm}{monash2}
        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
        \begin{textblock}{7}(12.7,27.9)\hfill
        \includegraphics[height=0.7cm]{AACSB}~~~
        \includegraphics[height=0.7cm]{EQUIS}~~~
        \includegraphics[height=0.7cm]{AMBA}
        \end{textblock}
        \vspace*{2cm}
        \begin{center}\Large
        Department of Econometrics and Business Statistics\\[.5cm]
        \footnotesize http://monash.edu/business/ebs/research/publications
        \end{center}\vspace{2cm}
        \begin{center}
        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
                \LARGE\@author\end{onehalfspace}
        }}
        \end{center}
        \vfill
                \begin{center}\Large
                \Month~\Year\\[1cm]
                Working Paper \@wp
        \end{center}\vspace*{2cm}}}
\def\pageone{{\sffamily\setstretch{1}%
        \thispagestyle{empty}%
        \vbox to \textheight{%
        \raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
        \vspace{2cm}\par
        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
        \hspace{1cm}\showjel\vss}}}
\def\blindtitle{{\sffamily
     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
        }}
\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}

\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
\let\maketitle\titlepage
\makeatother

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% HYPERLINKS
\RequirePackage{xcolor} % Needed for links
\definecolor{darkblue}{rgb}{0,0,.6}
\RequirePackage{url}

\makeatletter
\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
\makeatother
\hypersetup{
     citecolor=0 0 0,
     breaklinks=true,
     bookmarksopen=true,
     bookmarksnumbered=true,
     linkcolor=darkblue,
     urlcolor=blue,
     citecolor=darkblue,
     colorlinks=true}

%% KEYWORDS
\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}

%% ABSTRACT
\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
  {\end{minipage}}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[showonlyrefs]{mathtools}
\usepackage[no-weekday]{eukdate}

%% BIBLIOGRAPHY

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}

\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}

\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

\nojel

\RequirePackage[absolute,overlay]{textpos}
\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}
\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}

\nocover

\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
\addresses{\textbf{Fan Cheng}\newline
Monash University
\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
\textbf{Anastasios Panagiotelis}\newline
The University of Sydney
\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
\textbf{Rob J Hyndman}\newline
Monash University
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Cheng, Panagiotelis, Hyndman: \@date}
\makeatother

%% Any special functions or other packages can be loaded here.
\usepackage{tabu,placeins,algorithmic}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\urlstyle{tt}  % use monospace font for urls
\usepackage[cal = txupr]{mathalpha}
\geometry{margin=2.2cm}
\mathtoolsset{showonlyrefs}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
% \def\equationautorefname~#1\null{Equation~(#1)\null} % NOT showing Equation 1, but Section 2.1
\usepackage{bm}
\DeclareMathOperator\supp{supp}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

% Adjust headwidth in case user has changed geometry in header-includes
\renewcommand{\headwidth}{\textwidth}

\begin{document}
\maketitle
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
\end{keywords}

\newpage

\hypertarget{dckdeintro}{%
\section{Introduction}\label{dckdeintro}}

Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}. For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\). The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. Our objective in this paper is to propose KDE for data that lie on some unknown manifold.

When dealing with data that lie on a manifold, it is common to use dimension reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\). Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx} and \textcite{Cheng2021-dh}. One could easily apply KDE to the output vectors \(\pmb{y}\). However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,\]
where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\). By a change of variables, this would imply the following density for \(\pmb{y}\),
\begin{equation}
\label{eq:changevar}
Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
\end{equation}
where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\pmb{H}(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\). It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|\pmb{H}(\pmb{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant. Since \(\psi\) will typically be a non-linear function, the value of \(|\pmb{H}(\pmb{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data. Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm. \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space. \autoref{dckdeapplication} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage. Conclusions and discussions are presented in \autoref{dckdeconclusion}. Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.

\hypertarget{kderm}{%
\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}}

In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, which can be applied when the data lie on some known manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.

In the following, we denote \(M\) as the \(d\)-dimensional manifold from which our data are sampled. Points on this manifold are denoted \(\pmb{p}\) in general, with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample. Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this need not be the case. For instance, \(\pmb{p}_i\) may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each \(\pmb{p}_i\) only require some sense of distance between the `input' points, \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding, \(\pmb{y}_1,\dots,\pmb{y}_n\), where \(\pmb{y}_i\in\mathbb{R}^d\). We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\). Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(\|\pmb{y}\|)=K(0).
\end{align}
Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation. For instance, condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise. In our empirical section, more general kernel functions can be, and are, employed.

For data \(\pmb{y}_i\in\mathbb{R}^d\) with \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\pmb{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{y}\) is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
\end{equation}

The intuition behind this estimator is very clear for a uniform kernel. The density at a point \(\pmb{y}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\pmb{y}\), times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein. Kernel density estimators of this form can and have been applied directly on the output embedding \(\pmb{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.

\hypertarget{Pellet}{%
\subsection{Kernel Density estimation on manifolds}\label{Pellet}}

For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
\end{equation}

where \(d_g(\pmb{p}, \pmb{p}_i)\) denotes the geodesic distance between two points on the manifold \(\pmb{p}\) and \(\pmb{p}_i\) and \(\theta_{\pmb{p}_i}(\pmb{p})\) is known as the volume density function. The intuition behind the term \(K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)\) is relatively clear. For example, for a uniform kernel, the estimator at point \(\pmb{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\). However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance. An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold. A definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix. For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\pmb{p}\) is so large that some points `fall inside' the ball more than once. For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere. This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.

The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term. We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\). However, the volume of this ball must also be taken into account. In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center. The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.

More formally, the volume density function can be explained as follows. Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space, \(\pmb{v}\in T_{\pmb{p}}M\), to points on the manifold, \(\pmb{q}\in M\). Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and traveling along this geodesic at uniform speed \(\|\pmb{v}\|\) takes place in one unit of time. Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system. The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system. In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\pmb{H}(\pmb{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\pmb{p}_i\). For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.

\hypertarget{MetLearn}{%
\subsection{Riemannian metric estimation}\label{MetLearn}}

To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric. Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\). Using the coordinates from a dimension reduction algorithm \(\pmb{y}\), angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions. To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point. These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.

While full details are provided in \textcite{Perrault-Joncas2013-pq}, we briefly describe the Learn Metric algorithm here. There are four main steps in the algorithm. First, a weighted neighborhood graph is constructed, with edges between \(\pmb{p}_i\) and \(\pmb{p}_j\) when \(\pmb{p}_i\) is a K-nearest neighbor of \(\pmb{p}_j\) or vice versa, and edge weights depending on the distance between \(\pmb{p}_i\) and \(\pmb{p}_j\) on the manifold. Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph. Third, a dimension reduction method is applied to obtain the output embedding \(\pmb{y}_1,\dots,\pmb{y}_n\). Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator). Full details on these four steps are provided in \autoref{alg:learnmetric}. This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}. Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as suggested in the \emph{megaman} library.

As pointed out by \textcite{Perrault-Joncas2013-pq}, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\). In this case, the ranks of the matrices \(\pmb{H}_i\) are equal to \(d\). Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension \(d\) globally into \(d\)-dimensional Euclidean space. In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into \(d\)-dimesional Euclidean space. In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used (See Chapter 3 of \textcite{Lee2007-wq} for more about intrinsic dimension estimation). The \emph{R} library \emph{intRinsic} \autocite{Denti2021-qc} implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters EMBED }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;

>>>>

---
title: "Distortion corrected kernel density estimate on Riemannian manifolds"
blinded: 0 # Set to 1 for blinding
authors:
  - name: Fan Cheng
    affiliation: Monash University
    thanks: Fan.Cheng@monash.edu
  - name: Rob J Hyndman
    affiliation: Monash University
  - name: Anastasios Panagiotelis
    affiliation: The University of Sydney    
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D data from the Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
keywords:
  - manifold learning
  - variable bandwidth
  - Riemannian metric
  - geodesic distance
  - Gaussian kernels
bibliography: kderm.bib
output:
  bookdown::pdf_document2:
    template: asa_template.tex
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    dev: "pdf"
    keep_tex: yes
    includes:
      in_header: preamble.tex
# after_body: appendix.tex
link-citations: yes
---

```{r child="03-chap3-kderm.Rmd"}

```

```{r child="B-appB-kderm.Rmd"}

```
